---
title: "The ordinal number-line model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The ordinal number-line model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4.2, dpi = 96)
library(nlFit)
haveSGS <- requireNamespace("smartGridSearch2", quietly = TRUE)
```

## What the model does

The model simulates the estimation process one trial at a time. It does not
describe the shape of a participant's estimates. It produces the estimates, and
the shape is whatever comes out. Two processes act on every trial. The ordinal
process places the target between the two remembered numbers that bracket it and
uses rank alone, never distance. The linear process places the target at its own
value on the line. The response is a weighted average of the two placements, and
`accuracyPercent` is the weight on the linear process. Thompson and Cohen (2025,
*American Psychologist*, doi:10.1037/amp0001617) introduce the model and the
data it was built on.

The fitted parameters describe a participant, not a curve. They say where the
participant puts a number when nothing remembered brackets it, how many of the
participant's own earlier estimates are still available to anchor the next one,
how finely the participant separates two numbers near a landmark, and how much
weight the participant gives to responding accurately.

## Terms

A **visible reference point** is a labelled position the participant can see.
Pass these as `visibleReferencePoints`. They are required. A **conceptual
reference point** is a position the participant supplies without seeing it, such
as the middle of the line. Pass these as `conceptualReferencePoints`, and give
the probability that one is available on a run as `pConceptual`. A
**remembered estimate** is an estimate the participant made earlier in the same
run. It anchors later targets, but it is not a reference point. The distinction
matters in one place. A reference point is always discriminable from the target.
A remembered estimate may not be. The **bounds**, `lowerBound` and `upperBound`,
are the ends of the region the participant can respond in. No estimate is placed
outside them.

Two tasks use these arguments differently. In the bounded task the ends of the
line are labelled, so the bounds are also the visible reference points and the
same two numbers are passed twice. In the universal task the visible reference
points are the labelled values and the bounds are the edges of the screen.

## One run, one parameter at a time

The worked example is a universal number line. The screen runs from 0 to 100 and
carries two labels, 5 and 50. `targetOrder = "fixed"` takes the targets in the
order given, so each trial can be followed in turn.

```{r setup-line}
nl <- function(targets, ...) {
  ordinalNumberLine(targets = targets, upperBound = 100, lowerBound = 0,
                    visibleReferencePoints = c(5, 50),
                    targetOrder = "fixed", ...)
}
```

### The first estimate

Trial 1 presents 15. Nothing has been estimated yet, so the two anchors are the
label 5 and the label 50, and both are reference points. `firstEstimate` is the
proportion of that region at which the estimate lands. The value 15 plays no part
in the placement. Only the linear process knows the number.

The rule is not confined to trial 1. It applies whenever both anchors are
reference points, which happens again on trial 3. Target 60 lies above every
label, so it is bracketed by the label 50 and the upper screen edge, and it lands
at the same proportion of that region.

```{r first-estimate}
targets4 <- c(15, 6, 60, 10)
f.low  <- nl(targets4, firstEstimate = 0.2)
f.mid  <- nl(targets4, firstEstimate = 0.5)
f.high <- nl(targets4, firstEstimate = 0.8)
data.frame(trial = 1:4, target = targets4,
           f.0.2 = f.low$fEst, f.0.5 = f.mid$fEst, f.0.8 = f.high$fEst)
```

At `firstEstimate = 0.2` trial 1 puts 15 at `r f.low$fEst[1]`, which is one fifth
of the way from 5 to 50. At 0.8 it puts 15 at `r f.high$fEst[1]`. Trial 3 moves
with it, from `r f.low$fEst[3]` to `r f.high$fEst[3]`.

```{r first-estimate-plot, echo = FALSE}
plot(targets4, f.mid$fEst, type = "n", xlim = c(0, 100), ylim = c(0, 100),
     xlab = "target", ylab = "estimate")
abline(0, 1, col = "grey70")
abline(v = c(5, 50), col = "grey85")
points(targets4, f.low$fEst,  pch = 1)
points(targets4, f.mid$fEst,  pch = 16)
points(targets4, f.high$fEst, pch = 2)
legend("topleft", c("0.2", "0.5", "0.8"), pch = c(1, 16, 2),
       title = "firstEstimate", bty = "n")
```

### Ordinality and the memory window

Trial 2 presents 6. The remembered numbers are the labels 5 and 50, the screen
edges, and the estimate given to 15. Only the rank matters, so 6 falls between
the label 5 and the remembered 15, and it lands midway between the position of
the label and the position given to 15. Trial 4 presents 10, which falls between
the remembered 6 and the remembered 15 and lands midway between their estimates.
Every anchored target lands midway between its two anchors, so an error made
early is carried forward.

`memoryLength` is the number of previous estimates still available. With
`memoryLength = 1` only the estimate from the previous trial is remembered.

```{r memory}
m.full  <- nl(targets4, firstEstimate = 0.5, memoryLength = 4)
m.short <- nl(targets4, firstEstimate = 0.5, memoryLength = 1)
data.frame(trial = 1:4, target = targets4,
           memory.4 = m.full$fEst, memory.1 = m.short$fEst)
```

The first three trials are the same. On trial 4 the short window holds only the
estimate of 60, so nothing remembered brackets 10, both anchors are labels again,
and the first-estimate rule fires. Target 10 is placed at `r m.short$fEst[4]`, the
same place 15 was placed. A memory window shorter than the target set turns the
fitted function into a staircase.

```{r memory-plot, echo = FALSE}
plot(targets4, m.full$fEst, type = "n", xlim = c(0, 100), ylim = c(0, 100),
     xlab = "target", ylab = "estimate")
abline(0, 1, col = "grey70")
abline(v = c(5, 50), col = "grey85")
points(targets4, m.full$fEst,  pch = 16)
points(targets4, m.short$fEst, pch = 1)
legend("topleft", c("4", "1"), pch = c(16, 1), title = "memoryLength", bty = "n")
```

### Number sensitivity

Every number carries a range of equivalent numbers. The half width of that range
is `(1 - numberSensitivity)` times the distance to the nearest reference point.
Target 15 is 10 away from the label 5, so at `numberSensitivity = 0.8` its range
runs from 13 to 17. A later 14 falls inside that range and is treated as the same
number as 15. It is placed where 15 was placed, and the linear process aims at 15
rather than at 14.

```{r sensitivity}
s.fine   <- nl(c(15, 14), firstEstimate = 0.5, numberSensitivity = 1)
s.coarse <- nl(c(15, 14), firstEstimate = 0.5, numberSensitivity = 0.8)
data.frame(trial = 1:2, target = c(15, 14),
           s.1 = s.fine$fEst, s.0.8 = s.coarse$fEst)
```

At `numberSensitivity = 1` every range has zero width, and 14 gets its own
placement of `r s.fine$fEst[2]`. At 0.8 it takes 15's estimate of
`r s.coarse$fEst[2]`.

The test runs in both directions, so the pair collapses whichever number came
first.

```{r sensitivity-order}
nl(c(14, 15), firstEstimate = 0.5, numberSensitivity = 0.8)$fEst
```

Because the half width is scaled by the distance to the nearest reference point,
resolution is finest next to a label and coarsest midway between two labels.

```{r sensitivity-plot, echo = FALSE}
grid <- seq(0, 100, by = 0.5)
half <- 0.2 * pmin(abs(grid - 5), abs(grid - 50))
plot(grid, half, type = "l", xlab = "target",
     ylab = "half width of the range at s = 0.8")
abline(v = c(5, 50), col = "grey85")
```

### Accuracy

`accuracyPercent` is the weight the response gives to the linear process. At 0
the response is the ordinal placement. At 1 it is the value the linear process
aims at. Below is the same pair of trials at
`numberSensitivity = 0.8` and three weights.

```{r accuracy}
acc <- lapply(c(0, 0.5, 1), function(a) {
  nl(c(15, 14), firstEstimate = 0.5, numberSensitivity = 0.8, accuracyPercent = a)
})
data.frame(target = c(15, 14), a.0 = acc[[1]]$fEst,
           a.0.5 = acc[[2]]$fEst, a.1 = acc[[3]]$fEst)
```

Target 14 moves from `r acc[[1]]$fEst[2]` to `r acc[[3]]$fEst[2]` as the weight
rises. At full weight the estimate of 14 is `r acc[[3]]$fEst[2]`, not 14, because
sensitivity has already merged the two numbers and the linear process aims at the
number the participant thinks was presented.

```{r accuracy-plot, echo = FALSE}
est14 <- sapply(acc, function(x) x$fEst[2])
plot(NA, xlim = c(0, 50), ylim = c(0, 1), yaxt = "n", ylab = "",
     xlab = "position on the line")
abline(h = 0.5, col = "grey70")
points(c(5, 50), c(0.5, 0.5), pch = "|", cex = 1.6)
text(c(5, 50), 0.62, c("label 5", "label 50"), cex = 0.8)
points(est14, rep(0.5, 3), pch = 3, cex = 1.4)
text(est14, 0.38, paste("a =", c(0, 0.5, 1)), cex = 0.8)
```

### Conceptual reference points

A conceptual reference point is a reference point the participant may or may not
have on a given run. Each one is drawn independently once per run and is then
present on every trial of that run. `pConceptual` is the probability
of that draw. Here the participant may hold a conceptual point at 25.

```{r conceptual}
c.out <- nl(targets4, firstEstimate = 0.5,
            conceptualReferencePoints = 25, pConceptual = 0)
c.in  <- nl(targets4, firstEstimate = 0.5,
            conceptualReferencePoints = 25, pConceptual = 1)
data.frame(target = targets4, p.0 = c.out$fEst, p.1 = c.in$fEst)
```

The point moves the estimates two ways. It is an anchor, so it splits the region
between 5 and 50 in two, and it is a reference point, so it narrows the range of
equivalent numbers for every target near it.

```{r conceptual-plot, echo = FALSE}
plot(targets4, c.out$fEst, type = "n", xlim = c(0, 100), ylim = c(0, 100),
     xlab = "target", ylab = "estimate")
abline(0, 1, col = "grey70")
abline(v = c(5, 50), col = "grey85")
abline(v = 25, col = "grey85", lty = 2)
points(targets4, c.out$fEst, pch = 1)
points(targets4, c.in$fEst,  pch = 16)
legend("topleft", c("0", "1"), pch = c(1, 16),
       title = "pConceptual", bty = "n")
```

## Many runs

A participant sees the targets in an order the experimenter drew, and the model's
estimates depend on that order. `ordinalNumberLineSim()` runs the simulation
`loops` times, draws a fresh order on each run when `targetOrder = "random"`, and
returns the mean estimate. That mean is `fEst`, and it is the model's prediction
for a target. It is what a fit is scored against.

```{r sim}
set.seed(11)
parList <- list(upperBound = 100, lowerBound = 0,
                visibleReferencePoints = c(5, 50), firstEstimate = 0.5,
                accuracyPercent = 0.3, numberSensitivity = 0.8,
                targetOrder = "random")
ordinalNumberLineSim(c(6, 10, 15, 30, 45, 60, 15), parList, loops = 200)
```

Target 15 was presented twice, so it has two rows. `presentation` numbers the
occurrences of a target value in the order they were presented. The averaging is
by target value and presentation together. The first presentation of a value sits
earlier in a run on average than the second, so the two have different expected
estimates even under a random order. `loops` sets how many runs are averaged.
More loops give a more precise `fEst` and cost more time.

## Fitting a participant

A participant is a set of targets and the estimates they gave. Below is one
simulated from the model at known parameters with normal noise added, so the fit
can be checked against the truth.

```{r participant}
truth <- list(upperBound = 100, lowerBound = 0,
              visibleReferencePoints = c(5, 50), firstEstimate = 0.35,
              accuracyPercent = 0.45, numberSensitivity = 0.75,
              targetOrder = "random")
set.seed(2)
sim <- ordinalNumberLineSim(c(3, 6, 8, 10, 13, 15, 18, 22, 26, 30,
                              38, 45, 55, 62, 70, 80, 90),
                            truth, loops = 500)
noisy <- sim$fEst + rnorm(nrow(sim), 0, 3)
person <- data.frame(target = sim$target,
                     estimate = round(pmin(pmax(noisy, 0), 100), 1))
head(person)
```

`nlFit()` fits the model to one participant in a single call. `free` names the
parameters the search varies; every other parameter is held at the value given
through its own argument, so the line that runs a fit is also the record of what
was fitted and what was not. A parameter named in `free` is searched over its
whole domain, or over a range given as `free = list(numberSensitivity = c(0.5, 1))`.
A range may narrow a domain but not widen it.

The search is `smartGridSearch()`, which is in the smartGridSearch2 package and
is not on CRAN. `nlFit()` stops with the install line when it is missing, and the
chunks below are not run when it is missing here. `search = "quick"` and the small
`control` settings keep this vignette fast. A reported fit uses `search = "full"`,
which is the default, and its estimates are finer than the ones below.

```{r nlfit, eval = haveSGS, results = "hide", message = FALSE}
fit <- nlFit(person, upperBound = 100, lowerBound = 0,
             visibleReferencePoints = c(5, 50),
             free = c("firstEstimate", "accuracyPercent", "numberSensitivity"),
             search = "quick",
             control = list(multicore = FALSE, numLoops = 30, numIntervals = 12,
                            optParamListN = 5, optBoundLoops = 2,
                            loops = 40, finalLoops = 100))
```

Printing the fit gives the version it was made under, the parameters, which of
them were searched and over what range, and the fit statistics.

```{r nlfit-print, eval = haveSGS}
fit
```

Plotting it draws the participant's estimates against the targets, the identity
line, the visible reference points, and the fitted estimates.

```{r nlfit-plot, eval = haveSGS, fig.width = 5, fig.height = 5}
plot(fit)
```

The participant sits above the identity line at the low end and below it at the
high end, which is what an ordinal placement between two labels produces. The
values the search returns are not the values this participant was generated at.
The parameters trade off against one another, so on noisy estimates several sets
fit about equally well and the search returns the one that scores best, which
need not be the one the data came from.

`as.data.frame()` reduces the fit to one row: the five parameters and the fit
statistics.

```{r nlfit-row, eval = haveSGS}
as.data.frame(fit)
```

`nlFit()` writes nothing unless it is given a file name. Pass `sinkFilename` to
write a report recording the package version, the run, the number-line settings,
the parameter values and the fit, and `plotFileName` to write the plot to a pdf.

### What the search minimizes

`getOrdinalNumberlineFit()` runs the simulation at one set of parameters and
returns a single number, the statistic named by `minimizeStat`. That is the
objective `nlFit()` hands to the search, and it can be called on its own.
`pars.n` is the number of parameters the search varies, here three.

```{r objective}
getOrdinalNumberlineFit(person, upperBound = 100, lowerBound = 0,
                        visibleReferencePoints = c(5, 50),
                        firstEstimate = 0.35, accuracyPercent = 0.45,
                        numberSensitivity = 0.75, loops = 60, pars.n = 3)
```

`outputNLfitStats()` is the other end of the same sequence: it recomputes the fit
at one set of parameters and returns the fit statistics and the data with the
fitted values attached. `nlFit()` calls it at the best parameters the search
found, and the result is in the `fitted` element of the fitted object.

### Many participants

Fit one participant per call. Split the data by participant, fit each one, and
stack the one-row results.

```{r many, eval = FALSE}
fits <- lapply(split(allData, allData$participant), function(p) {
  nlFit(p, upperBound = 100, lowerBound = 0, visibleReferencePoints = c(5, 50))
})
do.call(rbind, lapply(fits, as.data.frame))
```

The data are fitted as they are given, so the shape of the dataframe decides what
kind of fit it is. One row per target is an averaged fit, which is what a dataset
with several trials per target has to be reduced to first.

```{r aggregate, eval = FALSE}
averaged <- aggregate(estimate ~ target, data = p, FUN = mean)
```

Leaving the trials as they are gives a trial-level fit instead. Add a
`presentation` column numbering the occurrences of each target value, and the fit
is matched to the simulation trial by trial rather than target by target. A
trial-level fit has more data points per participant at the same `pars.n`, so its
BIC is on a different scale. Do not compare it with the BIC of a fit to averaged
data.

## What changes across versions

Parameters keep their meaning. A future release will not redefine
`numberSensitivity` or `firstEstimate`, and an analysis run under a later version
will read the same way. Extensions arrive as new arguments whose defaults
reproduce the behaviour described here. Changes that do move the predictions are
listed in `NEWS.md`, and the version behind the published fits is kept under a
git tag.

## Acknowledgments

*This document was drafted with assistance from Claude (Anthropic). All technical content reflects the authors' design decisions and domain expertise.*
